---
title: "Introducing the nlshelper package"
author: "Remko Duursma"
date: 2017-04-24
categories: ["R"]
tags: ["plot", "regression", "nonlinear"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message= FALSE, warning=FALSE)
```


I recently organized a workshop on non-linear regression with R, covering the fundamentals and a few extras. When preparing the handout (which is [available here](http://www.hiercourse.com/nonlinear) as a PDF), I realized it is actually quite cumbersome to plot a fitted model with `nls`, together with the data. To make this process easier, I wrote a simple package - `nlshelper` - which does just that and a few other things. Read on for a simple demonstration, or visit the [Github repository](https://github.com/remkoduursma/nlshelper) here. The package is also on CRAN. 

# Plotting a single fitted curve

The main motivation for writing this package was to quickly be able to plot a fitted curve on top of the data. The `nlstools` package does already provide this functionality via `plotfit` but I found that function to be somewhat inflexible, and it also does not accept models fit with `nlsList` (more on that below).

The following example fits the Chapman-Richards equation to the `Loblolly` dataset, and plot the fitted curve with the data.

```{r, message=FALSE}
# Define function to fit
chapm <- function(x,Asym,b,c)Asym*(1-exp(-b*x))^c

# Fit model
# Note the specification of the starting values.
nls_lob <- nls(height ~ chapm(age, Asym, b,c),
               data=Loblolly,
               start=list(Asym=100, b=0.1, c=2.5))

# Plot fitted curve
library(nlshelper)
plot_nls(nls_lob, pch=16, points.col="cornflowerblue")
```

# Plotting a loess model

After having implemented the simple `plot_nls` function, it turned out that it can be used (or easily modified) for other objects, including those fit with `loess` or even non-linear quantile regression (`nlrq` from `quantreg`).


```{r}
library(nlshelper)
library(nlme)
chick <- as.data.frame(ChickWeight)
fit1 <- nlsList(weight ~ a*Time^b|Diet, data=chick, start=list(a=10, b=1.1))
plot_nls(fit1)
```

